Indicators for Ecosystem Condition in Norway

### 2024-09-27 gjengroingMetaModel.Rmd
Modified to just include data download script

The workflow spans two platforms including RStudio and Python.

## About the underlying data {#about-data-enc}

We rely on the following datasets:

-   [Nature type polygons](https://kartkatalog.miljodirektoratet.no/Dataset/Details/2031) are used to identify reference areas with good ecological condition.
-   NIBIO's [AR5](https://kartkatalog.geonorge.no/metadata/arealressurskart-fkb-ar5-arealtyper/280bbd7a-5ce9-4c83-9e15-ac162cabd8a6) is used as the population sample for each ecosystem type. The population sample are the areas used to determine ecological condition relative to the reference areas. We also use the AR5 data to identify forest surrounding the population polygons for defining the lower limit of ecological condition index (i.e.. poor condition),
-   LiDAR-derived digital elevation model from Kartverket's [høydedata](https://hoydedata.no/LaserInnsyn/). This includes both a terrain (DTM) and surface model (DSM). We calculate the canopy height model (DSM - DTM) to get the height of objects above the ground. From this we remove buildings, and what remains is vegetation - mostly trees but also some bushes or smaller woody plants.
-   Kartverket's [national 10m elevation model](https://kartkatalog.geonorge.no/metadata/dtm-10-terrengmodell-utm32/fd851873-f363-46f9-9fc6-bb1b403575df). We use this to stratify reference (good condition) and forest (poor condition) heights by elevational band. This is done in combination with bioclimatic zones - see next point.
-   Moen's [bioclimatic zones](https://artsdatabanken.no/Pages/181901/Bioklimatiske_soner). We use this to stratify reference (good condition) and forest (poor condition) heights by bioclimatic (also referred to as vegetation/climatic) zones.
-   [FKB building footprints](https://kartkatalog.geonorge.no/metadata/fkb-bygning/8b4304ea-4fb0-479c-a24d-fa225e2c6e97) are used to isolate vegetation in the LiDAR height data.
-   A [European satellite-based map](https://www.nature.com/articles/s41893-020-00609-y) of forest clear cuts is used for identifying AR5 forest patches in near-climax successional stages.
-   The [SSB 10km grid](https://kartkatalog.geonorge.no/metadata/statistisk-rutenett-5000m/32ac0653-d95c-446c-8558-bf9b79f4934e) is used for visualization purposes.
-   The regional delineation for Norway (five regions) are used for aggregating and reporting gjengroing condition values.

### Representativity in time and space {#rep-enc}

The index will cover the mainland of Norway. The analysis will be stratified by (1) åpne and (2) våtmark ecosystems. The former includes both semi-naturlig mark and naturlig åpne områder under skoggrensa (due to the fact that AR5 does not differentiate these), and the latter includes våtmark only. The LiDAR data cover a range of years and therefore the indicator represents conditions for circa 2010 to 2021.

### Original units {#units-enc}

The original units for ecological condition are meters. This is the height of the vegetation within referecne, polygon and forest areas.

### Temporal coverage {#temp-enc}

Circa 2010 to 2021. This is a single snapshot and not a change analysis. In the future, when LiDAR data has been repeated across the country, it may be possible to do a change assessment. In addition, the use of optical or radar satellite imagery may serve as a proxy for vegetation height using machine learning. This will allow for annual updates, depending on the uncertainty in the satellite-based maps of vegetation height.

### Additional comments about the dataset {#comm-enc}

None.

## Ecosystem characteristic

### Norwegian standard

Funksjonell sammensetning innen trofiske nivåer, reflecting the change in dominant growth form.

## Collinearities with other indicators

There is possibly a collinearity with the [primary production indicator](#NDVI-indicator-natopen) (primærproduksjon). The primary production indicator uses the normalized difference vegetation index (NDVI) as a proxy for vegetation production. NDVI can be correlated with vegetation height and consequently yield similar results to the LiDAR-based gjengroing indicator.

## Reference state and values

The methodology used to calculate the _gjengroing_ indicator is outlined in the schematic below. The workflow in the schematic is conducted for all reference and population polygons in Norway and repeated for each ecosystem type (åpne and våtmark), respectively. The indicator values are aggregated to a 50km grid and regional level at the end. The individual steps are discussed in turn in the following subsections.

```{r enc-workflow, fig.cap="Schematic illustration of how the encroachment indicator is calculated. The lower bound for poor condition is set by LiDAR heights from mature forest surrounding the population polygon. The upper bound for good condition is set by regional median LiDAR heights within NiN polygons in the same bioclimatic-elevation zone as the population polygon.", out.width='70%', echo=FALSE}
knitr::include_graphics("images/gjengroing_schematic.jpg")
```

### Reference state {#state-enc}

The reference state is defined as one where åpne ecosystems are indeed open, with little to no woody encroachment. 
For natural ecosystems this reflects a state where climate warming, wetland drainage, and alien species etc have not led to any net loss of these nature types and their internal functioning. 
For semi natural ecosystems it reflects a state where traditional husbandry is keeping he ecosystems open due to grazing and hay making.

### Reference values, thresholds for defining *good ecological condition*, minimum and/or maximum values {#ref-enc}
The NiN polygons including "Naturlig åpne områder under skoggrensa", "Semi-naturlig mark", and "Våtmark" with good ecological condition are used to define a reference gjengroing state. We use the aggregated "Tilstand" variable assigned to each NiN polygon. The 50th percentile of LiDAR-derived vegetation heights within these polygons is used to define the upper limit (ie. 1) of the scaled indicator value. We cannot define local reference values based on proximity, because the NiN polygons are spatially biased and not close to all AR5 population polygons. Therefore we calculate regional reference values using elevation (300m bands) and bioclimatic zones as stratification. We calculate the mean reference value for each unique combination of elevation-bioclimatic zone. When calculating the index for each population polygon, the reference value is inherited from the elevation-bioclimatic zone it falls within.

Once we have the reference vegetation height for a given ecosystem type and elevation-bioclimatic zone, we need to define the minimum (or worst/bad) condition. We use the 50th percentile of LiDAR-derived vegetation heights within AR5 skog polygons to define a climax vegetation successional stage where gjengroing is at its most extreme. In order not to include forest patches that have recently been harvested, we mask out any forest which has been clear-cut since 1986. Here, 1986 is a hard limit defined by the clear-cut dataset which was based on Landsat imagery. Therefore we can be assured that we are measuring forest that is at least 35 years old.

We use both local and regional "reference" approaches to defining poor ecological condition. To do this we calculate the forest heights in AR5 forest patches that fall within 200m of a population polygon (local reference). For population polygons that do not have forest within their proximity, we use the mean forest height within the unique elevation-bioblimatic zone the population falls within (regional reference).

To define the ecological condition of the population polygon, we measure 50th percentile of LiDAR-derived vegetation height. As a reminder, we use AR5 polygons that have types that are approximately compatible with the NiN ecosystem types. For våtmark we use the AR5 polygons defined as "Myr". For åpne ecosystems we use the AR5 polygons defined as either "Åpent fastmark" or "Innmarksbeite".

We acknowledge that the AR5 data set has several shortcomings in terms of lags in update frequency and with precision in the maps themselves. 
This indicator will therefore benefit strongly from new ecosystem delineation maps, that update more frequently, and preferably once that can differentiate between semi-natural and naturally open ecosystems.

The vegetation height percentiles are then scaled to between 0 and 1 using a sigmoid transformation [Oliver at al. (2021)](https://www.sciencedirect.com/science/article/pii/S1470160X21000066).

## Uncertainties {#err-enc}

For the indicator map at the polygon level there is no uncertainty associated with the indicator values. For aggregated indicator values (e.g. for 50km grid and regions), the uncertainty in the indicator value is calculated from the spatial variation in the polygon-level reference height values. We could use the alternative of calculating the spatial variation in the indicator values via bootstrapping using the EAtools package. However, given the extreme number of population polygons, the uncertainty values are extremely small and therefore we choose to use the standard deviation in reference vegetation heights, converted to the indicator scale.

## References

Links to data and resources are provided with hyperlinks in-line.

## Analyses {#analyses-enc}

### Data sets {#data-enc}

There are several datasets which are used in GEE which will not be imported into the R session here. These datasets have been obtained from the source and ingested into GEE by Zander Venter or Vegard Bakkestuen with the help of Miljødata section at NINA. They include

-   [AR5](https://kartkatalog.geonorge.no/metadata/arealressurskart-fkb-ar5-arealtyper/280bbd7a-5ce9-4c83-9e15-ac162cabd8a6)
-   [LiDAR-derived digital elevation model from høydedata](https://hoydedata.no/LaserInnsyn/).
-   [Kartverket's national 10m elevation model](https://kartkatalog.geonorge.no/metadata/dtm-10-terrengmodell-utm32/fd851873-f363-46f9-9fc6-bb1b403575df)
-   [FKB building footprints](https://kartkatalog.geonorge.no/metadata/fkb-bygning/8b4304ea-4fb0-479c-a24d-fa225e2c6e97)
-   [European forest clear-cut map](https://www.nature.com/articles/s41893-020-00609-y)

The remaining datasets will be imported into the R session.

@@

OK, now that we have our tiles, we can pass them to the canopy height model please see https://github.com/facebookresearch/HighResCanopyHeight for more details on the model.

######Zanders/Ida's script######
#### Regions {#reg-enc}

The regional delineation for Norway (five regions) are used for aggregating and reporting gjengroing condition values.

```{r}
regions <- sf::st_read("data/regions.shp", options = "ENCODING=UTF8") %>%
  mutate(region = factor(region))
```

#### SSB 10 and 50km grids

The [SSB 10km and 50km grids](https://kartkatalog.geonorge.no/metadata/statistisk-rutenett-5000m/32ac0653-d95c-446c-8558-bf9b79f4934e) are used for visualizing the distribution of available data and for aggregating gjengroing index scores.

```{r}
ssb10km <- st_read('/data/R/GeoSpatialData/Population_demography/Norway_SSB/Original/ssb_10km/ssb10km.shp')%>%
  # mutate the id value so that it aligns with the data coming from GEE
  mutate(SSBID = as.numeric(SSBID)/1000)
# Transform to the correct projection
ssb10km <- st_transform(ssb10km, st_crs(regions))

ssb50km <- st_read('/data/R/GeoSpatialData/Population_demography/Norway_SSB/Original/ssb_50km/ssb50km.shp')%>%
  # mutate the id value so that it aligns with the data coming from GEE
  mutate(SSBID = as.numeric(SSBID)/1000)
# Transform to the correct projection
ssb50km <- st_transform(ssb50km, st_crs(regions))
```

#### Bioclimatic regions

The [Moen's bioclimatic regions](https://data.artsdatabanken.no//Natur_i_Norge/Natursystem/Beskrivelsessystem/Regional_naturvariasjon/Bioklimatisk_sone) are imported from NINA R/GeoSpatialData/ server.

```{r}
bioclim <- st_read('/data/R/GeoSpatialData/BiogeographicalRegions/Norway_VegetationZones_Moen/Original/Vector/soner.shp') %>%
  mutate(NAVN = ifelse(NAVN == 'S°rboreal', 'Sørboreal', NAVN))
bioclim <- st_transform(bioclim, st_crs(regions))
```

#### NiN polygons

[NiN polygons](https://kartkatalog.miljodirektoratet.no/dataset/Details/2031) are imported here from the NINA servers:

```{r}
nin <- st_read('/data/R/GeoSpatialData/Habitats_biotopes/Norway_Miljodirektoratet_Naturtyper_nin/Original/versjon20221231/Natur_Naturtyper_NiN_norge_med_svalbard_25833/Natur_Naturtyper_NiN_norge_med_svalbard_25833.gdb')%>%
  # Simplify ecosystem names
  mutate(hovedøkosystem = recode(hovedøkosystem, 
                                 "Våtmark" = 'Våtmark',
                                 "Skog" = "Skog",
                                 "Ingen" = "Ingen",
                                 "Fjell" = "Fjell",
                                 "Semi-naturlig mark" = 'Semi-naturlig',
                                 "Naturlig åpne områder i lavlandet" = 'Naturlig åpne',
                                 "Naturlig åpne områder under skoggrensa" = 'Naturlig åpne')) %>% 
  # Fix any invalid geometries
  mutate(validGeo = st_is_valid(SHAPE)) %>%
  filter(validGeo) %>%
  # Filter out the ecosystems we are not interested in
  filter(!hovedøkosystem %in% c('Skog','Ingen','Fjell')) %>%
  # Get the main ecosystem codes in case we need them later
  mutate(hovedtype = substr(ninkartleggingsenheter, 4, 6),
         hovedtype = str_remove(hovedtype, '-'),
         id = identifikasjon_lokalid) %>%
  # Drop polygons with no condition score
  drop_na(tilstand) %>%
  dplyr::select(id, hovedøkosystem, hovedtype, naturtypekode, tilstand)
```

-   **Export the cleaned NiN data and upload to GEE before proceeding**

```{r}
# Only run if you are repeating this workflow for the first time
# nin %>%
#   dplyr::select(id, hovedøkosystem, tilstand) %>%
#   st_transform(st_crs(fylker_wgs84)) %>%
#   st_write('nin_cleaned.shp', append=FALSE)
```

-   **Now run the following GEE JavaScript code by clicking on the hyperlink:** [GEE script](https://code.earthengine.google.com/b2d06ad72c515fbbb3024f157987eedc) This script is also found in the SCR repository as "GEE_script.js". When you run the script, then click the Tasks tab in the GEE code editor and run the export tasks. The export files will appear in your Google Drive. Download the files from your Google Drive to the /data/gjengroing/From_GEE directory. This script extracts three CSV files:
    -   'area_cover_grid.csv' - this gives the area coverage of NiN, AR5 and LiDAR data for each 10x10km SSB grid cell.
    -   'elevation_stratified.shp' - this is a 300m elevation band stratification that is simply used for visualizing the strata here in R.
    -   'vegHeights.zip' - this is a folder with hundreds of CSV files. These files contain the vegetation heights for reference, population, and forest polygons over Norway. The exports were split by 50km grid to prevent hitting user memory limits in GEE. Unzip the folder before you upload to ./DATA/From_GEE/

#### Data from Google Earth Engine

Once you have run the GEE script above, and downloaded the data to the `/gjengroing/R_project/DATA/From_GEE/` directory, you can proceed with this R workflow.

Create functions to import vegetation height files and stratify data by elevation band and vegetation/climate zone

```{r}
# Function to read in multiple files resulting from GEE exports
readVegHeightFiles <- function(uniqueString){
  dir <- '/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/gjengroing/R_project/DATA/From_GEE/vegHeights/'
  files <- list.files(dir)[str_detect(list.files(dir), uniqueString)]
  #print(files)

  dat <- tibble()
  for (i in files){
    dat <- dat %>% 
      bind_rows(read_csv(paste0(dir, i)) %>%
                  mutate(ssbid = substr(str_split(i, '_')[[1]][3], 1, 14)))
  }
  return (dat)
}

# Vegetation type lookup
vegLookup <- tibble(
  vegClimZone = c(1,2,3,4,5,6), 
  vegClimZoneLab = c('Nemoral','Boreonemoral','Sørboreal','Mellomboreal','Nordboreal','Alpin')
  )

# Elevation stratification and vegetation type cleaning function
cleanElevVegType <- function(data){

  dataOut <- data %>%
    mutate(e = elevation) %>%
    mutate(elevation = ifelse(e < 300, '0-300',
                              ifelse(e > 300 & e < 600, '300-600',
                                     ifelse(e > 600 & e < 900, '600-900',
                                     ifelse(e > 900 & e < 1200, '900-1200', '>1200'))))) %>%
    mutate(vegClimZone = round(vegClimZone)) %>%
    left_join(vegLookup, by = 'vegClimZone') %>%
    dplyr::select(-e, -vegClimZone) %>%
    drop_na(vegClimZoneLab, elevation)

  return (dataOut)

}

```

Import the CSV files generated in GEE related to area covers and elevation bands:

```{r}
# Import data coverages per SSB 10km grid cell
areaCovers <- read_csv('/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/gjengroing/R_project/DATA/From_GEE/area_cover_grid.csv') %>%
  dplyr::select(-'.geo', -'system:index')

# Calculate percentage cover relative to the land area in each cell
areaCoversPerc <- areaCovers %>%
  gather(key, val, aapne_ar5, vaatmark_ar5,aapne_nin, vaatmark_nin, lidarCover, skog) %>%
  mutate(type = ifelse(str_detect(key, 'ar5'), 'ar5', 
                       ifelse(str_detect(key, 'nin'), 'nin', 
                              ifelse(key == 'lidarCover', 'lidar', NA)))) %>%
  mutate(ecosystem = ifelse(str_detect(key, 'aapne'), 'Semi-/Naturlig åpne', 
                       ifelse(str_detect(key, 'vaatmark'), 'Våtmark', NA))) %>%
  mutate(areaPerc = val/land*100,
         areaPerc = ifelse(is.na(areaPerc), 0, areaPerc)) %>%
  dplyr::select(SSBID, key, type,ecosystem, areaPerc)

elevStrata <- st_read('/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/gjengroing/R_project/DATA/From_GEE/elevation_stratified.shp') %>%
  mutate(elevBand = ifelse(mode == 1, '0-300',
                           ifelse(mode == 2, '300-600',
                                  ifelse(mode == 3, '600-900',
                                         ifelse(mode == 4, '900-1200', '>1200')))),
         elevBand = factor(elevBand, levels=c('0-300','300-600','600-900','900-1200','>1200')))
elevStrata <- st_transform(elevStrata, st_crs(regions))

```

Import the CSV files generated in GEE of vegetation heights (circa 15 min runtime):

```{r}

if (runFromScratch){

  # Import median LiDAR heights for NiN polygons
  refVaatmarkRaw <- readVegHeightFiles('vaatmark_ref') %>%
    mutate(ref = chm) %>%
    dplyr::select(-'.geo', -'system:index', -chm)
  refAapneRaw <- readVegHeightFiles('aapne_ref') %>%
    mutate(ref = chm) %>%
    dplyr::select(-'.geo', -'system:index', -chm)

  # Add elevation and veg/clim zones and summarise per strata
    # this will form the upper limit of the ecological condition score (ie. good condition)
  refVaatmark <- cleanElevVegType(refVaatmarkRaw) %>%
    group_by(elevation, vegClimZoneLab) %>%
    summarise(ref = mean(ref, na.rm=T))
  refAapne <- cleanElevVegType(refAapneRaw) %>%
    group_by(elevation, vegClimZoneLab) %>%
    summarise(ref = mean(ref, na.rm=T))

  # Export to file for reference
  refVaatmark %>%
    st_write('data/encroachment/refVaatmark.csv')
  refAapne %>%
    st_write('data/encroachment/refAapne.csv')

  # Import median LiDAR heights for AR5 polygons for population measurement
  popVaatmarkHt <- readVegHeightFiles('vaatmark_pop')  %>%
    mutate(pop = chm)%>%
    dplyr::select(-chm)
  popAapneHt <- readVegHeightFiles('aapne_pop')  %>%
    mutate(pop = chm)%>%
    dplyr::select(-chm)

  # Turn into spatial objects using the .geo column in the CSV
  popVaatmarkHt <- st_as_sf(bind_cols(popVaatmarkHt,geojsonsf::geojson_sf(popVaatmarkHt$.geo)))
  popAapneHt <- st_as_sf(bind_cols(popAapneHt,geojsonsf::geojson_sf(popAapneHt$.geo)))

  # Import median LiDAR heights for forest (skog) within 200m of population polygons
    # this will form the lower limit of the ecological condition score (ie. poor condition)
  popVaatmarkSkog <- readVegHeightFiles('vaatmark_skog') %>%
    mutate(skog = chm)%>%
    dplyr::select(-chm, -'.geo', -ssbid)
  popAapneSkog <- readVegHeightFiles('aapne_skog') %>%
    mutate(skog = chm)%>%
    dplyr::select(-chm, -'.geo', -ssbid)

  # Combine the population heights with the forest heights
  popVaatmark <- popVaatmarkHt %>%
    left_join(popVaatmarkSkog, by = 'system:index')%>%
    dplyr::select(-'.geo')
  popAapne <- popAapneHt %>%
    left_join(popAapneSkog, by = 'system:index')%>%
    dplyr::select(-'.geo')

  # Add elevation and veg/clim zones to the combined population and forest heights
  popVaatmark <- cleanElevVegType(popVaatmark)
  popAapne <- cleanElevVegType(popAapne)

  # Get regional reference values for forest height
    # this is to fill in for population polygons which did not have any forest within 200m
  vaatmarkSkogRegionalRef <- popVaatmark %>%
    as_tibble() %>%
    group_by(elevation, vegClimZoneLab) %>%
    summarise(skogFill = mean(skog, na.rm=TRUE)) %>%
    drop_na(skogFill)
  aapneSkogRegionalRef <- popAapne %>%
    as_tibble() %>%
    group_by(elevation, vegClimZoneLab) %>%
    summarise(skogFill = mean(skog, na.rm=TRUE)) %>%
    drop_na(skogFill)

  # Combine all the above into one data frame
  #set.seed(123)
  vaatmarkHts <- popVaatmark %>%
    #sample_n(100000) %>%
    as_tibble() %>% dplyr::select(-geometry) %>%
    left_join(refVaatmark, by=c('vegClimZoneLab', 'elevation')) %>%
    gather(type, height, ref, pop, skog) %>%
    group_by(`system:index`, elevation, vegClimZoneLab, type) %>%
    summarise(height = mean(height)) %>%
    pivot_wider(values_from = height, names_from = type) %>%
    # If population height is less than the reference value, then it is automatically "good" condition - ie. it inherits the reference height so that the rescaled value will be 1
    mutate(pop = ifelse(pop < ref, ref, pop)) %>%
    drop_na(pop, ref) %>%
    left_join(vaatmarkSkogRegionalRef, by = c('vegClimZoneLab', 'elevation')) %>%
    mutate(skog = ifelse(is.na(skog), skogFill, skog)) %>%
    dplyr::select(-skogFill) %>%
    filter(skog < 30)
  #colSums(is.na(vaatmarkHts))

  aapneHts <- popAapne %>%
    #sample_n(100000) %>%
    as_tibble() %>% dplyr::select(-geometry) %>%
    left_join(refAapne, by=c('vegClimZoneLab', 'elevation')) %>%
    gather(type, height, ref, pop, skog) %>%
    group_by(`system:index`, elevation, vegClimZoneLab, type) %>%
    summarise(height = mean(height)) %>%
    pivot_wider(values_from = height, names_from = type) %>%
    # If population height is less than the reference value, then it is automatically "good" condition - ie. it inherits the reference height so that the rescaled index will be 1
    mutate(pop = ifelse(pop < ref, ref, pop)) %>%
    drop_na(pop, ref) %>%
    left_join(aapneSkogRegionalRef, by = c('vegClimZoneLab', 'elevation')) %>%
    mutate(skog = ifelse(is.na(skog), skogFill, skog)) %>%
    dplyr::select(-skogFill) %>%
    filter(skog < 30)
  #colSums(is.na(aapneHts))

} else {

  refVaatmark  <- read_csv('data/encroachment/refVaatmark.csv')
  refAapne <- read_csv('data/encroachment/refAapne.csv')

}

```

```{r ref-open-enc, fig.cap="Refenerence vegetation height in semi-natural and naturally open ecosystems under the forest line in Norway."}
refAapne %>%
  mutate(elevation = factor(elevation, levels = c("0-300", "300-600", "600-900", "900-1200", ">1200")),
         vegClimZoneLab = factor(vegClimZoneLab)) %>%
  ggplot(aes(x = elevation, y = ref))+
  geom_bar(stat="identity")+
  theme_bw(base_size = 12)+
  theme(axis.text.x = element_text(angle = -45, vjust = 0.5, hjust = 0.2, size = 12))+
  facet_wrap(.~vegClimZoneLab)+
  labs(y = "Reference vegetation height (m)",
       x = "Elevation (meters above sea level)")
```

```{r ref-wet-enc, fig.cap="Refenerence vegetation height in wetland ecosystems under the forest line in Norway."}
refVaatmark %>%
  mutate(elevation = factor(elevation, levels = c("0-300", "300-600", "600-900", "900-1200", ">1200")),
         vegClimZoneLab = factor(vegClimZoneLab)) %>%
  ggplot(aes(x = elevation, y = ref))+
  geom_bar(stat="identity")+
  theme_bw(base_size = 12)+
  theme(axis.text.x = element_text(angle = -45, vjust = 0.5, hjust = 0.2, size = 12))+
  facet_wrap(.~vegClimZoneLab)+
  labs(y = "Reference vegetation height (m)",
       x = "Elevation (meters above sea level)")
```

### Explore data coverages

Here we will make maps of NiN, AR5 and LiDAR coverage over Norway to get a feel for the data distributions. These areas were calculated in GEE and exported to CSV in the GEE script provided above. All we need to do now is join them with the SSB 10km grid and visualize.

```{r coverage-map-enc, fig.height=5, fig.width=15, fig.cap="Map of Norway showing coverage of different data sets."}

c1 <- ssb10km %>% 
  left_join(areaCoversPerc %>%
              filter(type == 'nin') %>%
              group_by(SSBID) %>%
              summarise(areaPerc = sum(areaPerc, na.rm=T)))  %>%
  ggplot() +
  geom_sf(aes(fill=areaPerc), color=NA) +
  geom_sf(data=regions, fill=NA, size=0.3) +
  scale_fill_gradientn(colours = c('black', 'orange'),
                       name = '% cover',
                       limits=c(0, 2),
                       oob=scales::squish) +
  theme_void() +
  theme(legend.position = c(0.7, 0.4)) +
  ggtitle('A) NiN polygons with good condition')
#c1
c2 <- ssb10km %>% 
  left_join(areaCoversPerc %>%
              filter(type == 'ar5') %>%
              group_by(SSBID) %>%
              summarise(areaPerc = sum(areaPerc, na.rm=T)))  %>%
  ggplot() +
  geom_sf(aes(fill=areaPerc), color=NA) +
  geom_sf(data=regions, fill=NA, size=0.3) +
  scale_fill_gradientn(colours = c('black', 'orange'),
                       name = '% cover',
                       limits=c(0, 100),
                       oob=scales::squish) +
  theme_void() +
  theme(legend.position = c(0.7, 0.4)) +
  ggtitle('B) AR5 polygons')
#c2
c3 <- ssb10km %>% 
  left_join(areaCoversPerc %>%
              filter(type == 'lidar') %>%
              group_by(SSBID) %>%
              summarise(areaPerc = sum(areaPerc, na.rm=T)))  %>%
  ggplot() +
  geom_sf(aes(fill=areaPerc), color=NA) +
  geom_sf(data=regions, fill=NA, size=0.3) +
  scale_fill_gradientn(colours = c('black', 'orange'),
                       name = '% cover',
                       limits=c(0, 100),
                       oob=scales::squish) +
  theme_void() +
  theme(legend.position = c(0.7, 0.4)) +
  ggtitle('C) LiDAR canopy height model')
#c3

coverageFig <- grid.arrange(c1, c2, c3, ncol=3, widths=c(1,1,1), padding = unit(0, "line"), newpage = T)

```

### Bioclimatic zone and elevation strata for regional reference values

We will visualize the elevation bands and bioclimatic zones that are used to define regional reference values for poor (forest vegetation height) and good (NiN vegetation height) ecological condition. For each unique spatial combination of these strata (n = 21), we calculate a mean reference value later on in the analysis. Note that poor ecological condition is preferably calculated using a local reference (forest height within 200m of population polygon), but a regional reference is used where there are no nearby forest patches.

```{r sones-and-bands, fig.cap="Bioclimatic sones and elevation bands used to define strata in the enchroachment indicator."}

bioclimPlot <- bioclim %>%
  ggplot() + 
  geom_sf(aes(fill = NAVN), color=NA)+
    scale_fill_manual(values=brewer.pal(10, 'Paired'), name = 'Zone')+
    ggtitle('A) Bioclimatic zones') +
    theme_void()

elevPlot <- elevStrata %>%
  ggplot() +
  geom_sf(aes(fill = elevBand), color=NA, size=0.001)+
    scale_fill_manual(values=brewer.pal(5, 'Pastel1'), name = 'HASL (M)')+
    ggtitle('B) Elevation bands')+
    theme_void()

strataPlot <- grid.arrange(bioclimPlot, elevPlot, ncol=2, widths=c(1,1), padding = unit(0, "line"), newpage = T)

```

We can see that while AR5 polygons are well distributed across the country, the NiN polygons are clustered and biased toward low elevation regions of the country. The LiDAR data covers nearly the entire country with some exceptions in the north. Future iterations of this work will need to attempt to quantify the biases introduced by the relative distributions of reference and population polygons and by the datasets used to quantify gjengroing (in this case LiDAR).

### Scaled indicator values at polygon level (circa 120 min runtime)

We will calculate the gjengroing index for each population polygon over Norway using a Sigmoid scaling function.

```{r}

# Define Sigmoid scaling function
scaleSigmoid <- function(variable){
  refLow <- min(variable)
  refHigh <- max(variable)
  thr <- (refHigh-refLow)/2
  indicator_LowHigh <- (variable - refLow)/(refHigh - refLow)
  indicator_LowHigh[indicator_LowHigh[] < 0] <- 0
  indicator_LowHigh[indicator_LowHigh[] > 1] <- 1
  indicator_sigmoid <- 100.68*(1-exp(-5*(indicator_LowHigh)^2.5))/100
  return (round(indicator_sigmoid, 4))
}

if (runFromScratch){

  # Calcualte gjengroing index values
  vaatmarkIndexPoly <- vaatmarkHts  %>%
    # Gather into long format so taht we can mutate the scaleSigmoid() function
    gather(type, height, ref, pop, skog) %>%
    group_by(`system:index`) %>%
    mutate(index = scaleSigmoid(height)) %>%
    # Because short vegetation is good condition, we must invert the scaled indicator
    mutate(index = 1-index) %>%
    # Get the indicator values for the population polygons only (ref and skog will be 1 and 0)
    filter(type == 'pop') %>%
    dplyr::select(-type, -height) %>%
    # Join with original data to get the actual median vegetaiton heights for each indicator score
    left_join(vaatmarkHts %>% dplyr::select("system:index", pop, ref, skog)) %>%
    left_join(popVaatmark %>% dplyr::select("system:index", ssbid))

  # do the same for åpne polygons - will not comment this because same as above
  aapneIndexPoly <- aapneHts  %>%
    gather(type, height, ref, pop, skog) %>%
    group_by(`system:index`) %>%
    mutate(index = scaleSigmoid(height)) %>%
    mutate(index = 1-index) %>%
    filter(type == 'pop') %>%
    dplyr::select(-type, -height) %>%
    left_join(aapneHts %>% dplyr::select("system:index", pop, ref, skog)) %>%
    left_join(popAapne %>% dplyr::select("system:index", ssbid))

  # Make into spatial objects
  vaatmarkIndexPolySpat <- popVaatmark %>%
    dplyr::select("system:index") %>%
    left_join(vaatmarkIndexPoly) %>%
    filter(!is.na(index))
  aapneIndexPolySpat <- popAapne %>%
    dplyr::select("system:index") %>%
    left_join(aapneIndexPoly) %>%
    filter(!is.na(index))

  # Transform to correct projection
  vaatmarkIndexPolySpat <- st_transform(vaatmarkIndexPolySpat, st_crs(regions))
  aapneIndexPolySpat <- st_transform(aapneIndexPolySpat, st_crs(regions))

} else {

  vaatmarkIndexPolySpat <- st_read('/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/gjengroing/R_project/OUTPUT/vaatmark_index.shp')%>%
    mutate(`system:index` = id,
           elevation = elev, 
           vegClimZoneLab = vegZn) %>%
    dplyr::select(-id, -elev, -vegZn)
  aapneIndexPolySpat <- st_read('/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/gjengroing/R_project/OUTPUT/aapne_index.shp')%>%
    mutate(`system:index` = id,
           elevation = elev, 
           vegClimZoneLab = vegZn) %>%
    dplyr::select(-id, -elev, -vegZn)

}

```

Here we will visualize the spatial variation in scaled gjengroings indicator values, as well as the vegetation heights for reference (good condition), indicator, and mature forest (poor condition) over the country. We could plot all the polygons, but it is computationally intensive and very difficult to visualize spatial variations over the entire country. Therefore we will calculate the area-weighted mean in values for each bioclimatic-elevation strata (n = 21) over the country.

```{r multiple-maps-enc, fig.width=10, fig.cap="Maps of scaled and unscaled indicator values, as well as reference values. For the unscaled variables the legend is truncated at 5 m to ensure the same colour gradient for all maps."}

# Create intersection of the two strata layers
#elevStrata <- st_transform(elevStrata, st_crs(regions))
bioClimElev <- bioclim %>%
  st_intersection(elevStrata) %>%
  mutate(elevation = elevBand,
         vegClimZoneLab = NAVN)%>%
  dplyr::select(elevation, vegClimZoneLab) %>%
  rowid_to_column("ID")

if (runFromScratch){

  # Get area-weighted mean for reference, forest and population vegetation heights 
  # for each unique bioclimatic-elevation zone combination
  vaatmarkIndexStrata <- bioClimElev %>% 
    #sample_n(10000) %>%
    st_join(vaatmarkIndexPolySpat %>% 
                mutate(area = as.numeric(st_area(geometry))) ,st_intersects) %>%
    group_by(ID) %>%
    summarise_at(vars(pop, ref, skog, index), funs(weighted.mean(., w=area, na.rm=TRUE)))%>% 
                as_tibble() %>% 
                dplyr::select(-geometry)

  aapneIndexStrata <- bioClimElev %>% 
    #sample_n(10000) %>%
    st_join(aapneIndexPolySpat %>% 
                mutate(area = as.numeric(st_area(geometry))) ,st_intersects) %>%
    group_by(ID) %>%
    summarise_at(vars(pop, ref, skog, index), funs(weighted.mean(., w=area, na.rm=TRUE)))%>% 
                as_tibble() %>% 
                dplyr::select(-geometry)

  # Write out for import later
  vaatmarkIndexStrata %>% write_csv('/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/gjengroing/R_project/DATA/vaatmarkIndexStrata.csv')
  aapneIndexStrata %>% write_csv('/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/gjengroing/R_project/DATA/aapneIndexStrata.csv')

} else {

  vaatmarkIndexStrata <- read_csv('/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/gjengroing/R_project/DATA/vaatmarkIndexStrata.csv')
  aapneIndexStrata <- read_csv('/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/gjengroing/R_project/DATA/aapneIndexStrata.csv')

}

# Make plots of indicator heights and scaled condition values for each strata polygon

theme_legend_1 <- theme(legend.key.size = unit(0.25, 'cm'), #change legend key size
                        legend.key.height = unit(0.25, 'cm'), #change legend key height
                        legend.key.width = unit(0.25, 'cm'), #change legend key width
                        legend.title = element_text(size=6), #change legend title font size
                        legend.text = element_text(size=6)) #change legend text font size

makeStrataHeightMap <- function(data, var, title, limits){
  p <- bioClimElev %>%
    left_join(data, by = c('ID'))  %>%
    mutate(response = .[[var]]) %>%
    ggplot() + 
    geom_sf(aes(fill = response), color=NA)+
      scale_fill_gradientn(colors=viridis::viridis(10),
                           limits = limits,
                           oob = scales::squish,
                           name = 'Veg ht (m)')+
      ggtitle(title) +
      theme_void() +
    theme(plot.title = element_text(size=9)) +
    theme_legend_1

  return (p)
}

>>>>

This script downloads image tiles to be used in the meta canopy height model (INSERT CROSSREF).

@@

OK, now that we have our tiles, we can pass them to the canopy height model please see https://github.com/facebookresearch/HighResCanopyHeight for more details on the model.
Please follow (INSERT CROSSREF)
